---
title: "Species summary report"
author: "Dan Carpenter"
date: "03/01/2022"
params: "Species"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load libraries
library(readr)
library(readxl)
library(dplyr)
library(lubridate)
library(janitor)
library(sf)
library(tmap)
library(tmaptools)
#library(OpenStreetMap)
library(leaflet)
library(ggplot2)
library(rnrfa)

# read data

records <- read_csv("data/myrecords.csv")
sac <- st_read("data/SACs.shp")

```

## Introduction

This report provides a summary of the species recorded by Dan Carpenter, as well as a summary of the Special Areas of Conservation around Oxford.

## Species records

The following section provides a summary of species recorded in 2020 and 2021, largely during the lockdowns imposed to control the spread of coronavirus.

```{r cleaning}

# clean names
records <- janitor::clean_names(records)

# convert dates
records$date <- lubridate::dmy(records$date)

```

A summary table of the species groups Dan recorded is presented below.  In total, 186 records were recorded across 13 species groups.

```{r group_summary}

grp <- records %>%
  group_by(species_group) %>% 
  count(name = "grp_n") %>% 
  ungroup() %>% 
  select(species_group, grp_n) %>% 
  janitor::adorn_totals()

knitr::kable(grp, col.names = c("Species group", "No. of records"))

```

The most recorded species are shown in the graph below.

```{r freq_spp}

spp2 <- records %>% 
  group_by(common_name, species) %>% 
  count() %>% 
  ungroup() %>% 
  arrange(desc(n)) %>% 
  select(common_name, species, n)

d <- spp2 %>% 
  filter(n > 3) %>% 
  ggplot(aes(x=common_name, fill=common_name, y=n)) +
  geom_col()+
  theme_light() +
  theme(axis.text.x=element_text(angle=90,hjust=1)) +
  ggtitle("Most recorded species") +
  labs(x = "Common name", y = "No. of records", fill = "Common name")
d

```

Below is an interactive webmap of the species records.

```{r spp_webmap}

# convert grid refs to E,N

x <- osg_parse(records$map_ref)

# write to data

records$east <- x$easting
records$north <- x$northing

# create wkt from eastings and northings

records$wkt <- paste0("POINT(", records$east, " ", records$north, ")")


# convert to sf

rs <- st_as_sf(records, wkt="wkt")
class(rs)
rs

# set crs
st_crs(rs) <- 27700
rs

# transform data for leaflet
rs <- st_transform(rs, 4326)

# create leaflet map

m <- leaflet(rs) %>% 
  addTiles() %>% 
  addMarkers(popup = paste0("Species: ", rs$common_name,"<br>",
                           "(", rs$species, ")"))
m

```

## Special Areas of Conservation

Below is a summary of Special Areas of Conservation (SACs) around Oxford.

```{r}

sacd <- as.data.frame(sac)

sacd <- sac %>% 
  mutate(area_ha = area/10000)


```

The mean area of the SACS is `r round(mean(sacd$area_ha), 2)` hectares, with a range of `r round(range(sacd$area_ha), 2)` hectares.  The largest SAC has an area of `r round(max(sacd$area_ha), 2)` hectares and the smallest `r round(min(sacd$area_ha), 2)` hectares.

A map of the SACs is presented below.

```{r sac_map}

sac <- st_transform(sac, 4326)

# read OSM raster data
#osm <- tmaptools::read_osm(sac, ext=1.1)

tm <- tm_shape(sac) + 
      tm_fill("blue") + 
      tm_borders("blue")
tm

```

